\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{stmaryrd}    % \llbracket, \rrbracket
\usepackage{mathpartir}  % inference rules
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{microtype}

\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{example}{Example}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\state}{\Sigma}
\newcommand{\reagent}[2]{\mathsf{Reagent}(#1, #2)}
\newcommand{\run}[2]{\mathsf{run}(#1, #2)}
\newcommand{\abortt}{\!\Uparrow\!}         % abort marker
\newcommand{\committ}{\!\Downarrow\!}      % commit marker
\newcommand{\sq}{\mathbin{\texttt{>{>}}}}
\newcommand{\choice}{\mathbin{+}}

\title{A High-Level Correctness Condition for Reagent Execution}
\author{Working note -  seeds for meeting with Prof. Kartik}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We propose a correctness condition for the concurrent execution of reagents
à la Turon~\cite{TuronThesis,TuronICFP12}, building on the transactional
memory semantics tradition of Guerraoui--Kapalka~\cite{Opacity},
Doherty--Groves--Luchangco--Moir--Scott~\cite{TMS}, and, in particular,
Attiya--Gotsman's client-observable framework~\cite{AttiyaGotsman}. The
condition generalises \emph{opacity} to account for two features of reagents
that ordinary STM does not have: external choice via $(+)$ and synchronous
rendezvous via channel swap. We state the condition as a history-level
predicate and sketch what it obliges of the layered implementations in
\texttt{01\_kcas}--\texttt{06\_reagent\_cps}.
\end{abstract}

%-----------------------------------------------------------------------------
\section{Introduction}

Reagents are composable atomic actions. A reagent $r : \reagent{A}{B}$
transforms an input of type $A$ into an output of type $B$ while effecting an
atomic state transition on shared memory and, possibly, on synchronous
channels. Reagents compose in three ways:
\begin{itemize}
  \item \emph{sequencing} $r_1 \sq r_2$: the two reagents commit as one
    atomic step;
  \item \emph{choice} $r_1 \choice r_2$: an external (non-deterministic)
    choice between two alternatives, only one of which is committed;
  \item \emph{rendezvous} $\mathsf{swap}\;ep$: a two-party synchronous
    action pairing with a dual swap on the opposite endpoint.
\end{itemize}

Classical STM correctness conditions --- opacity in particular --- pin down
the intuition that ``each committed transaction appears to execute
instantaneously''. They do not, out of the box, cover (i) the fact that a
reagent with a $(+)$ at its root has a choice of \emph{which} committed
action the linearisation sees, nor (ii) the fact that a swap's linearisation
point coincides, by construction, with that of its dual.

We aim to state a single condition, \emph{reagent opacity}, that captures
both. The condition is history-based and can be checked without reference to
any particular implementation; in the layered codebase it yields concrete
proof obligations for each layer.

%-----------------------------------------------------------------------------
\section{Background: Transactional Opacity}

\paragraph{Histories.} Following Herlihy--Wing and the
Attiya--Gotsman~\cite{AttiyaGotsman} presentation, a \emph{history} $H$ is a
finite sequence of \emph{events} of the form $\langle p, e \rangle$, where
$p$ is a thread identifier and $e$ is one of
\[
  \text{inv}(r, v), \quad \text{res}(r, w), \quad \text{ret}(r),
\]
respectively invoking reagent $r$ with input $v$, responding with output
$w$, or aborting/retrying before any commit. Invocations are paired with
responses or aborts in program order within each thread. The projection of
$H$ to thread $p$ is written $H | p$.

\paragraph{Sequential histories.} A history is \emph{sequential} if no two
invocations are pending concurrently: every invocation is immediately
followed by its matching response or abort.

\paragraph{Reagent semantics.} We assume a denotational semantics
$\sem{\cdot}$ mapping reagents to partial functions on a state space
$\state$:
\[
  \sem{r} : \state \times A \rightharpoonup \state \times B.
\]
$\sem{r}$ is partial because a reagent may \emph{retry} (its precondition is
not met at the given state and input). For the primitive reagents:
\begin{align*}
  \sem{\mathsf{lift}\;f}(\sigma, a)            &= (\sigma, f(a)) \\
  \sem{\mathsf{upd}\;\ell\;f}(\sigma, b)       &= (\sigma[\ell \mapsto v'], c) \text{ if } f(\sigma(\ell), b) = \mathsf{Some}(v',c),\ \text{undefined otherwise} \\
  \sem{r_1 \sq r_2}(\sigma, a)                 &= \sem{r_2}(\sigma', b) \text{ where } (\sigma', b) = \sem{r_1}(\sigma, a) \\
  \sem{r_1 \choice r_2}(\sigma, a)             &\in \{\sem{r_1}(\sigma, a)\} \cup \{\sem{r_2}(\sigma, a)\}\ \text{if either is defined}
\end{align*}

For $\mathsf{swap}$ we defer to \S\ref{sec:rendezvous}.

\paragraph{Opacity (Guerraoui--Kapalka).} A history $H$ of STM transactions
is \emph{opaque} iff there exists a sequential history $H_s$ equivalent to
$H$ such that:
\begin{enumerate}
  \item $H_s$ respects the real-time order of completed transactions in $H$
        (if $t_1$'s response precedes $t_2$'s invocation in $H$, then $t_1
        \prec t_2$ in $H_s$);
  \item $H_s$ is \emph{legal}: each transaction reads values consistent with
        running the prefix of committed transactions sequentially;
  \item aborted transactions in $H_s$ also see consistent snapshots (i.e.\
        they never observe an intermediate, non-sequential state).
\end{enumerate}

Clause 3 is what distinguishes opacity from strict serialisability; it is
what prevents an aborted transaction from observing ``impossible'' states
that could, e.g., cause an infinite loop or segmentation fault in user code.

%-----------------------------------------------------------------------------
\section{Reagent Opacity}

We generalise opacity along two axes.

\subsection{Choice}

For $r = r_1 \choice r_2$, a committed response event $\text{res}(r, w)$ is
justified if \emph{exactly one} of $r_1, r_2$ produces $w$ at the
linearisation point. Aborted branches must not be observable: if $r_1$ was
tried first and then abandoned, the linearisation must not show writes from
$r_1$ interleaved with committed transactions.

This is automatic under Herlihy--Wing linearisability once we say the
linearisation step for $r_1 \choice r_2$ is \emph{the step of whichever
branch committed}. The work is in the implementation, not the specification.

\subsection{Rendezvous}\label{sec:rendezvous}

Swap is not a unary transaction: it requires a \emph{partner}. Denotationally
we model a swap pair as a single joint step:
\[
  \sem{\mathsf{swap}\;ep \; \mid \; \mathsf{swap}\;\overline{ep}}
  (\sigma, (a, c)) = (\sigma, (c, a)),
\]
where $\overline{ep}$ is $ep$'s dual. (For swaps embedded inside larger
reagents $r \sq \mathsf{swap}\;ep \sq r'$, we treat $r \sq r'$ and its dual
chain as a joint atomic step --- this is the point of reagents.)

A swap appears in the history as two events $\text{inv}(r_1, a)$ and
$\text{inv}(r_2, c)$ from different threads. In the linearisation these two
events must be \emph{co-linearised}: they occupy a single combined step,
with outputs $\text{res}(r_1, c)$ and $\text{res}(r_2, a)$ emerging from that
single step.

\subsection{The Condition}

\begin{definition}[Reagent Opacity]
  Let $H$ be a history of reagent invocations. $H$ is \emph{reagent-opaque}
  iff there exists a sequential history $H_s$ and a co-linearisation map
  $\mu$ on swap events such that:
  \begin{enumerate}
    \item $H_s$ respects real-time order among completed non-swap events.
    \item For each pair of swap events $e_1, e_2$ with $\mu(e_1) = e_2$,
          $e_1$ and $e_2$ are co-linearised: they are adjacent in $H_s$
          and form a single atomic step.
    \item $H_s$ is legal: for each committed $\run{r}{v} \Rightarrow w$ in
          $H_s$ with pre-state $\sigma$ and post-state $\sigma'$,
          $(\sigma', w) \in \sem{r}(\sigma, v)$.
    \item For each $r$ of the form $r_1 \choice r_2$ with committed
          response $w$ in $H_s$, the transition $(\sigma, v) \to (\sigma',
          w)$ equals $\sem{r_1}(\sigma, v)$ or $\sem{r_2}(\sigma, v)$
          (and only one).
    \item Every aborted/retried event in $H$ observes a state $\sigma_a$
          that is reachable as an intermediate state in $H_s$: i.e., $H_s$
          extended with a virtual aborted transaction reading $\sigma_a$
          is still legal.
  \end{enumerate}
\end{definition}

Clause 5 is opacity's safety-of-aborts property. Clauses 2 and 4 are
reagent-specific.

\begin{remark}[Relation to linearisability]
  Reagent opacity strictly refines the (library-local) linearisability
  condition on reagent operations, since it additionally constrains aborts
  (clause 5) and forbids intermediate states from leaking through retries.
  Where only committed reactions are observed, reagent opacity degenerates
  to strict serialisability plus the rendezvous and choice constraints.
\end{remark}

\begin{remark}[Relation to Attiya--Gotsman]
  The Attiya--Gotsman framework for STM consistency conditions defines
  client-observable equivalences between an implementation and a sequential
  specification. Reagent opacity slots into that framework: the
  \emph{sequential specification} is given by $\sem{\cdot}$ with a rule
  pairing swap invocations; the \emph{implementation} is any concrete
  reagent runtime (our layered \texttt{01\_kcas}--\texttt{06\_reagent\_cps});
  and \emph{correctness} is the existence of a linearisation witnessing
  reagent opacity. The reagent-specific clauses appear as side conditions
  on the linearisation map, not as substrate changes.
\end{remark}

%-----------------------------------------------------------------------------
\section{Proof Obligations for the Layered Implementations}

The condition breaks into per-layer obligations.

\paragraph{\texttt{01\_kcas} (GKMZ k-CAS).} Each $\mathsf{Kcas.atomically}$
call is a single atomic step on its locations. This is a classical
linearisability obligation for k-CAS, discharged by the two-phase protocol
(descriptor install; status flip).

\paragraph{\texttt{03\_stm} (\texttt{Xt}).} The log-based STM on top of
k-CAS induces \emph{ordinary} opacity for compositions of
$\mathsf{upd}$, $\mathsf{read}$, $\mathsf{cas}$, and $\sq$. The log's
\texttt{CMP} entries supply the read-validation required by clauses 3 and
5; \texttt{Xt.or\_else}'s log snapshot restores clause 4's ``only one
branch observable'' property.

\paragraph{\texttt{05\_lightweight} (direct-style \texttt{swap}).} The
direct-style channel swap lies \emph{outside} the \texttt{Xt} log. A history
in which a reagent $\mathsf{swap}\;ep \sq \mathsf{cas}\;\ell\;v\;v'$ commits
is \textbf{not} in general reagent-opaque under this definition: the swap
linearises at the rendezvous (when the partner is fulfilled), but the cas
linearises later at $\mathsf{Xt.commit}$, and an intervening thread can
observe the gap. This mismatch is the motivation for the CPS rewrite.

\paragraph{\texttt{06\_reagent\_cps} (CPS with \texttt{try\_swap}).}
Reagent opacity holds for compositions not involving swap, and for
$\mathsf{try\_swap} \choice r$ at the \emph{branch-selection} level
(clause~4) when no partner is present. It does not hold for
$\mathsf{try\_swap} \sq r$ when a partner has been fulfilled but $r$ then
retries: in that case the partner observes the rendezvous (clause~2 gives
them a completed swap event) while our thread retries, violating clause~3.

A faithful implementation would require the offer/reservation protocol
described in~\cite{TuronThesis} \S5: the offer slot's Pending$\to$Fulfilled
CAS is part of the same atomic bundle as the downstream reagent's CASes,
so the partner's linearisation point and ours coincide exactly as demanded
by clause~2.

%-----------------------------------------------------------------------------
\section{Open Questions}

\begin{itemize}
  \item \textbf{Fairness.} The condition above is safety-only. It says
    nothing about liveness: a pair of swap invocations may wait forever for
    a partner without violating reagent opacity. Is there a liveness
    companion? Weak progress (some reagent always makes progress) is
    probably the right level.
  \item \textbf{Nested choice.} Can clause~4 be strengthened to account for
    $(r_1 \choice r_2) \sq r_3$? The sequence should commit with
    $r_3$-in-the-same-atomic-step, regardless of which branch of the choice
    won. The condition above handles this through the denotation $\sem{r_1
    \sq r_3} \cup \sem{r_2 \sq r_3}$, but a more direct characterisation is
    desirable.
  \item \textbf{Retries and starvation.} In the minimal CPS implementation,
    a reagent returning \texttt{ShouldRetry} loops with \texttt{Sched.yield}.
    An opacity-preserving implementation should also rule out livelock
    under adversarial scheduling.
  \item \textbf{Relation to TMS1/TMS2.} Reagent opacity can likely be
    expressed as an instance of TMS-style spec with custom atomic steps
    (joint swap, branch projection). Working this out in detail would
    pin the condition into an established framework.
\end{itemize}

%-----------------------------------------------------------------------------
\begin{thebibliography}{99}

\bibitem{TuronThesis}
  A.~Turon.
  \emph{Understanding and expressing scalable concurrency}.
  PhD thesis, Northeastern University, 2013.

\bibitem{TuronICFP12}
  A.~Turon.
  Reagents: expressing and composing fine-grained concurrency.
  \emph{ICFP 2012}.

\bibitem{Opacity}
  R.~Guerraoui and M.~Kapalka.
  On the correctness of transactional memory.
  \emph{PPoPP 2008}.

\bibitem{TMS}
  S.~Doherty, L.~Groves, V.~Luchangco, M.~Moir, M.~Scott.
  Towards formally specifying and verifying transactional memory.
  \emph{Formal Aspects of Computing}, 2013.

\bibitem{AttiyaGotsman}
  H.~Attiya and A.~Gotsman.
  Transactional memory: models and algorithms (and client-observable
  consistency conditions more broadly).
  See: Koskinen--Parkinson, Lesani--Palsberg, and the Attiya--Hans--Gotsman
  line on consistency-condition refinements.

\end{thebibliography}

\end{document}
